---
title: "Mesh generation for INLA model"
author: "Morera-Pujol, Virginia"
date: "June 2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_knit$set(echo = TRUE)
```


## Preparation

### Load map of ireland to use as boundary, and point data to check they're inside the boundary

```{r load data, echo = T, message=FALSE, warning = FALSE}
ireland <- st_read("data/ireland_ITM.shp")
all_data <- read.csv("data/all_data.csv", row.names = NULL)

all_data_sf <- all_data %>% 
  st_as_sf(coords = c("Longitude", "Latitude")) %>% 
  st_set_crs(st_crs(ireland)) %>% 
  st_transform(crs = projKM)

ireland <- ireland %>% 
  st_transform(crs = projKM)
```


### Simplify and buffer 

- Simplify with a tolerance of 10 km 
- Buffering 15 km from the coast to avoid having points on the edge

```{r inner boundary, echo = T, message=FALSE, warning = FALSE}
boundary <- ireland %>% 
  mutate(Country = "Ireland") %>% 
  group_by(Country) %>%
  summarise(geometry = sf::st_union(geometry)) %>%
  ungroup() %>% 
  st_simplify(dTolerance = 10) %>% 
  st_buffer(dist = 15)
```

- Buffer the inner boundary by 200 km more to create outer boundary

```{r outer boundary, echo = T, message=FALSE, warning = FALSE}
boundary2 <- boundary %>% 
  st_buffer(dist = 200)
```

### Plot to check what we've got

```{r plot1, echo = T, message=FALSE, warning = FALSE}
ggplot(boundary) +
  geom_sf(data = boundary2, col = "red", fill = NA) +
  geom_sf(fill =NA) + 
  geom_sf(data = ireland, fill = NA, col = "darkgray") + 
  # geom_sf(data = all_data_sf, aes(col = Deer.Presence)) + 
  theme_bw()          
```

## Creation of mesh

### Transform to sp objects

```{r transform, echo = T, message=FALSE, warning = FALSE}
boundary_sp <- as_Spatial(boundary)

boundary2_sp <- as_Spatial(boundary2)

all_data_sp <- all_data_sf %>% 
  as_Spatial()

crs(boundary_sp)
crs(boundary2_sp)
crs(all_data_sp)
```

### Create boundary objects

- Transform inner boundary into inla segment object
- Transform outer boundary into inla segment object

```{r boundaries, echo = T, message=FALSE, warning = FALSE}
boundary_in <- inla.sp2segment(boundary_sp)
boundary_out <- inla.sp2segment(boundary2_sp)

boundary_in$crs
boundary_out$crs
```

### Create mesh

- Cutoff of 10 km (segments can't be smaller than 10 km, that's why we needed the simplification)
- The edge of the triangles in the inner mesh is 20km
- The edge of the triangles in the outer mesh is 50km 

```{r creating mesh, echo = T, message=FALSE, warning = FALSE}
mesh <- inla.mesh.2d(boundary = list(boundary_in, boundary_out), 
                     max.edge = c(20, 100), 
                     cutoff = 20, crs = boundary_out$crs)

mesh$crs

inla.identical.CRS(mesh$crs, boundary_sp@proj4string)
```


### Create larger mesh to test effects

- Cutoff of 20 km (segments can't be smaller than 20 km, that's why we needed the simplification)
- The edge of the triangles in the inner mesh is 40km (max dimension of map/10)
- The edge of the triangles in the outer mesh is 80km (double the inner edge) 

```{r creating mesh, echo = T, message=FALSE, warning = FALSE}
meshVLarge <- inla.mesh.2d(boundary = list(boundary_in, boundary_out), 
                          max.edge = c(60, 120), 
                          cutoff = 20, crs = boundary_out$crs)

meshLarge <- inla.mesh.2d(boundary = list(boundary_in, boundary_out), 
                          max.edge = c(40, 110), 
                          cutoff = 20, crs = boundary_out$crs)

inla.identical.CRS(meshLarge$crs, boundary_sp@proj4string)
inla.identical.CRS(meshVLarge$crs, boundary_sp@proj4string)

```


### Plot mesh

And add points to check all of them are within the inner boundary
```{r plotting mesh, echo = T, message=FALSE, warning = FALSE, fig.width = 7*0.8, fig.height = 10*0.7}
plot(meshLarge)
# plot(all_data_sp, add = T, col = "red")
plot(boundary_sp, border = "green", add = T)
plot(boundary2_sp, border = "orange", add = T)

par(mfrow = c(2,2))
plot(mesh)
plot(meshLarge)
plot(meshVLarge)
par(mfrow = c(1,1))
```

## Creation of mask

This will allow us to hide the predictions on the outer boundary from our results (probably there's a better way but just in case)
This code is taken from Haakon Bakka's barrier model from the website https://haakonbakka.bitbucket.io/btopic107.html#4_barrier_models and it runs all at once

### Code to create the mask

```{r creating mask, echo = T, message=FALSE, warning = FALSE}
tl = length(mesh$graph$tv[,1]) # Number of triangles of the mesh
posTri = matrix(0, tl, 2) # matrix containing the coordinates of each triangle's 

# we fill it with the vertex "coordinates here
for (t in 1:tl) {
  # Take the vertex of triangles
  temp = mesh$loc[mesh$graph$tv[t, ], ]
  # Compute barycenter which sintetize the triangle
  posTri[t,] = colMeans(temp)[c(1,2)]
}

#transform it in Spatial Points
posTri = SpatialPoints(posTri)
posTri@proj4string <- boundary_out$crs

normal = over(boundary_sp, posTri, returnList = T) # this are the polygons contained in the inner boundary, the "good" ones

normal = unlist(normal)
barrier.triangles = setdiff(1:tl, normal) # this are the polygons of the outer boundary, the ones we want to hide

# build a polygon that contains all polygons that are outer boundary and not inner. 
poly.barrier = inla.barrier.polygon(mesh, barrier.triangles)
```

### Plot the mask to check everything is ok

```{r plot2, echo = T, message=FALSE, warning = FALSE, fig.width = 7*0.8, fig.height = 10*0.7}
plot(mesh)
plot(poly.barrier, col = "gray", add = T)
```


```{r saving, eval = FALSE, echo=T, message=FALSE, warning=FALSE}
saveRDS(poly.barrier, file = "data/barrier.RDS")
saveRDS(mesh, file = "data/mesh.RDS")
saveRDS(meshLarge, file = "data/meshLarge.RDS")
saveRDS(meshVLarge, file = "data/meshVLarge.RDS")
saveRDS(boundary2_sp, file = "data/outer_boundary.RDS")
saveRDS(boundary_sp, file = "data/inner_boundary.RDS")
```
